---
title: "TADs Relaxation upon induced RAD21 Degradation in HCT116 Cells "
author: "Theodore Busby/Gianluca Pegoraro"
date: "January, 21st 2022"
output: html_document
---

```{r load-libraries}
library(tidyverse)
library(SpatialTools)
library(fs)
library(reshape2)
library(data.table)
library(ggthemes)
library(ggpointdensity)
source("R/Plotters.R")
```

### Image acquisition and analysis conditions

-   Microscope: CV7000
-   Objective: 60X water
-   Binning: 1X1
-   Image Analysis Software: HiTIPS v?

```{r set-options, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  fig.path = "output/",
  cache = FALSE,
  cache.path = "cache/",
  dev = c("png"),
  fig.width = 15,
  fig.height = 7,
  message = FALSE,
  warning = FALSE
)

theme_set(theme_bw())
theme_update(
  axis.text.x = element_text(
    angle = -45,
    hjust = 0,
    vjust = 0.5
  ),
  axis.text.y = element_text(hjust = 0)
)
```

Specify x,y pixel size (microns) and z-step (microns).

```{r set-resolution}
xy_res <- 0.108
z_step <- 0.5
```

### Set Experimental Treatments metadata

Read and process the experimental metadata layout file

```{r plate-layout}
treat_levs <- c("None", "Vehicle", "Auxin_30",
                "Auxin_60", "Auxin_180", "Auxin_360")

plate_layout <- read_tsv("metadata/plate_layout.txt") %>%
  filter(!is.na(treat)) %>%
  mutate(treat = factor(treat, levels = treat_levs))

glimpse(plate_layout)
```

```{r treatment-layout, echo=FALSE}
plot_plate(table = plate_layout,
           property = treat,
           discrete_property = T,
           legend = "Experimental\nTreatment",
           title = "Experimental Layout")
```

```{r probeset-layout, echo=FALSE}
plot_plate(table = plate_layout,
           property = probe,
           discrete_property = T,
           legend = "Probe Set",
           title = "FISH Probe Set Layout")
```

### Read Object Level Data

Set `glob` patterns for directory searches for cell level data and spot data,resepectively. The spot level data is stored in a single .csv file per well where all the spots in all the channels are stored. The `channel` variable value indicates the channel in which the spot was detected according to this mapping:

-   3: Red (561)
-   4: Far Red (640)

```{r set-glob}
glob_c <- "data*/well_nuclei_results/*.csv" # Pattern for Single Cell data files
glob_s <- "data*/well_spots_locations/*.csv" # Pattern for Spot level data files
```

Read and process the cell-level data. Filter nuclei that are irregularly shaped (`solidity`) and that are too small (`area`).

```{r read-cells}
cell_tbl <- dir_info("data", 
       recurse = T, 
       glob = glob_c) %>%
  filter(size > "1K") %>%
  rowwise() %>%
  summarise(fread(path)) %>%
  rename(cell_index = label,
         x = `centroid-0`,
         y = `centroid-1`) %>%
  select(Experiment, column, row, field_index,
         time_point, cell_index, x:solidity) %>%
  arrange(Experiment, column, row, field_index,
          time_point, cell_index) %>%
  filter(area >= 30) 
  
glimpse(cell_tbl)
```

Read and process the spot level data. Convert the x, y and z coordinates to microns (They were originally calculated in pixels) and filter spots that do not belong to any nucleus. These have a `cell_index` value of `0`.

```{r read-spots}
spot_tbl <- dir_info("data", 
       recurse = T, 
       glob = glob_s) %>%
  filter(size > "1K") %>%
  rowwise() %>%
  summarise(fread(path)) %>%
  rename(spot_index = V1,
         x = x_location,
         y = y_location,
         z = z_location) %>%
  #conversion from pixels to microns
  mutate(across(x:y, list(mic = ~ .x * xy_res)),
         z_mic = z * z_step) %>%
  select(column, row, field_index,
         time_point, cell_index, spot_index, 
         channel, x:z_mic) %>%
  #filter spots not in the nucleus
  filter(cell_index != 0) %>%
  arrange(column, row, field_index,
          time_point, cell_index, channel)

glimpse(spot_tbl)
```

Use the `spot_tbl` to calculate the number of spots per cell in each color.

```{r calc-n-spots}
cell_n_spots_tbl <- spot_tbl %>%
  group_by(column, row, field_index,
           time_point, cell_index, channel) %>%
  count() %>%
  pivot_wider(names_from = channel,
              names_prefix = "n_spots_channel_",
              values_from = n,
              values_fill = 0)

glimpse(cell_n_spots_tbl)
```

```{r n-red-spots, echo=F}
cell_n_spots_tbl %>%
  ggplot(aes(x = n_spots_channel_3,
             y = ..density..)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 0:5) +
  coord_cartesian(xlim = c(-1, 5)) +
  facet_grid(vars(row), vars(column)) +
  xlab("Number of Red Spots per Cell") +
  ggtitle("Red Spots") 
```

```{r n-farred-spots, echo=F}
cell_n_spots_tbl %>%
  ggplot(aes(x = n_spots_channel_4,
             y = ..density..)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 0:5) +
  coord_cartesian(xlim = c(-1, 5)) +
  facet_grid(vars(row), vars(column)) +
  xlab("Number of Far Red Spots per Cell") +
  ggtitle("Far Red Spots") 
```

### Spot Filtering

Filter cells that have 2 spots or more in both the 561 and 640 channels.

```{r}
cell_n_spots_filt_tbl <- cell_n_spots_tbl %>%
  semi_join(cell_tbl, by = c("row", "column", "field_index",
                             "time_point", "cell_index")) %>%
          filter(n_spots_channel_3 > 1 & n_spots_channel_3 == n_spots_channel_3)

glimpse(cell_n_spots_filt_tbl)
```

Filter out spots that do not belong to the correct population of cells obtained in the previous chunk.

```{r}
spot_filt_tbl <- spot_tbl %>% 
  semi_join(cell_n_spots_filt_tbl, 
            on = c("column", "row", "field_index",
                   "time_point", "cell_index")) %>%
  left_join(plate_layout, by = c("row", "column")) %>%
  data.table()
  
glimpse(spot_filt_tbl)
```

### Red/FarRed Spot Distances Calculations

Calculate 2D spot distances.

```{r calc-spot-dist, results="hide"}
#Green-Red
fr_dist_2D <- spot_filt_tbl[channel %in% 3:4, # Only  Red and FarRed spots
                              reshape2::melt(dist2(as.matrix(.SD[channel == 3,
                                                       list(x_mic,y_mic)]),
                                         as.matrix(.SD[channel == 4,
                                                       list(x_mic,y_mic)])),
                                   value.name = "fr_dist"),
                          by = list(row, 
                                    column,
                                    treat,
                                    probe,
                                    field_index, 
                                    time_point, 
                                    cell_index)]

setnames(fr_dist_2D, c("Var1","Var2"), c("r_index","f_index"))


glimpse(fr_dist_2D)
```

Red-FarRed proximity calculations. **Important: this calculations find the closest Red/FarRed pairs (by calculating the the minimum distance) on a per Red Spot basis**.

```{r proximityR-G, warning=F}
setkey(fr_dist_2D, row, column, treat, probe,
       field_index, time_point, cell_index, r_index)

fr_dist_min_2D <- fr_dist_2D[, .SD[which.min(fr_dist),], 
                        by = key(fr_dist_2D)]
```

### Single Allele Distances Distributions

Generate density plots on per Treatment and Probe basis (Pooled wells).

```{r pooled-sgRNA-gr-density, echo=F}
fr_dist_min_2D %>%
  ggplot(aes(x = fr_dist,
             color = treat)) +
  geom_density() +
  scale_color_viridis_d(name = "Treatment") +
  coord_cartesian(xlim = c(0, 4)) +
  facet_wrap(vars(probe)) +
  xlab("2D Red/Green Distance (microns)") +
  ylab("Density")
``` 

Generate density plots on per Treatment and Probe basis (Independent wells).

```{r single-sgRNA-fr-density, echo=F}
fr_dist_min_2D %>%
  ggplot(aes(x = fr_dist,
             color = treat,
             group = interaction(row, column))) +
  geom_density() +
  scale_color_viridis_d(name = "Treatment") +
  facet_wrap(vars(probe)) +
  coord_cartesian(xlim = c(0, 4)) +
  xlab("2D Red/FarRed Distance (microns)")
```

Document the information about the analysis session

```{r sessionInfo, include=TRUE, echo=TRUE, results="markup"}
sessionInfo()
```
