---
title: "FISH 2D Distance/210521 Cohesin Complex sgRNA Miniscreen"
author: "Theodore Busby/Gianluca Pegoraro"
date: "June, 22 2021"
output: html_document
---

```{r load-libraries}
library(tidyverse)
library(SpatialTools)
library(fs)
library(reshape2)
library(data.table)
```

### Experimental conditions
- Cells: HCT116-Cas9 Clone A
- Transfection reagent: RNAiMAX, XYZ nanoliters per well
- sgRNA: Synthego Knockout Kit v2, 2 nM final
- Cells fixed 72 hrs post transfection 
- DNA FISH:
  - Green Probe: XYZ
  - Red Probe: WQK

### Image acquisition and analysis conditions
- Microscope: CV7000
- Objective: 60X water
- Binning: 1X1
- Image Analysis Software: Columbus 2.9.1

```{r set-options, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  fig.path = "output/",
  cache = FALSE,
  cache.path = "cache/",
  dev = c("png"),
  fig.width = 15,
  fig.height = 7,
  message = FALSE,
  warning = FALSE
)

theme_set(theme_bw())
theme_update(
  axis.text.x = element_text(
    angle = -45,
    hjust = 0,
    vjust = 0.5
  ),
  axis.text.y = element_text(hjust = 0)
)
```

### Define Metadata
Design a tibble of sgRNA treatments based on `metadata/Mini Screen Plate layout`
```{r}
sgrna_v = c("PDS5B", "SMC1A", "NIPBL",
            "STAG2", "STAG1", "SMC3",
            "RAD21", "CTCF", "PDS5A",
            "WAPL", "MAU2", "SCRAMBLE",
            "PLK1", "SCRAMBLE", "PLK1")

metadata <- tibble(row = rep(2:16, 4),
                   col = rep(3:6, each = 15),
                   sgrna = rep(sgrna_v, 4)) %>%
            mutate(well = paste0(LETTERS[row], col))
```


### Read Object Level Data
Set `glob` patterns for directory searches for cell level data and spot data,respectively.

- Green (488)
- Red (561)

```{r set-glob}
glob_c <- "data/*Nuclei Final[0].txt" # Pattern for Single Cell data files
glob_g <- "data/*Spots 488[1].txt" # Pattern for Green Spot level data files
glob_r <- "data/*Spots 561[2].txt" # Pattern for Red Spot level data files
```

```{r unpack-packaged-data, include=FALSE}
has_miniscreen_inputs <- function(data_dir = "data") {
  if (!dir.exists(data_dir)) {
    return(FALSE)
  }
  files <- list.files(data_dir, recursive = TRUE, full.names = FALSE)
  all(
    any(grepl("Nuclei Final\\[0\\].*\\.txt$", files)),
    any(grepl("Spots 488\\[1\\].*\\.txt$", files)),
    any(grepl("Spots 561\\[2\\].*\\.txt$", files))
  )
}

archive_path <- "data_raw/miniscreen_columbus_exports.zip"
if (!has_miniscreen_inputs() && file.exists(archive_path)) {
  dir_create("data")
  unzip(archive_path, exdir = "data")
}
```

Read  and process the cell-level data. Filter nuclei that are irregularly shaped (`solidity`) and that are too small (`area`).
```{r read-cells}
nuclei <- dir_info("data", 
       recurse = TRUE, 
       glob = glob_c) %>%
  filter(size > "1K") %>%
  rowwise() %>%
  summarise(fread(path)) %>%
  select(screen = ScreenName,
         plate = PlateName,
         well = WellName,
         row = Row,
         col = Column,
         field = Field,
         timepoint = Timepoint,
         nuclei_index = `Nuclei Final - Object No in Nuclei`,
         nuclei_selected_index = `Nuclei Final - Object No in Nuclei Selected`,
         nuclei_final_index = `Object Number`,
         n_green_spots = `Nuclei Final - Number of Spots 488- per Cell`,
         n_red_spots = `Nuclei Final - Number of Spots 561- per Cell`,
         nuc_area = `Nuclei Final - Nucleus Area [µm²]`) %>%
  left_join(metadata, on = c("row", "col", "well")) %>%
  # filter out row A, where the FISH did not work
  filter(!is.na(sgrna))
  
glimpse(nuclei)
```
Read and process the green spot level data.
```{r read-green-spots}
green_spots <- dir_info("data", 
       recurse = T, 
       glob = glob_g) %>%
  filter(size > "1K") %>%
  rowwise() %>%
  summarise(fread(path)) %>%
  select(screen = ScreenName,
         plate = PlateName,
         well = WellName,
         row = Row,
         col = Column,
         field = Field,
         timepoint = Timepoint,
         spot_index = `Object Number`,
         nuclei_index = `Spots 488 - Object No in Nuclei population`,
         x = `Position X [µm]`,
         y = `Position Y [µm]`,
         z = `Spots 488 - Z-plane Mean`) %>%
  # Assign color label
  mutate(color = "green")

glimpse(green_spots)
```
Do the same for the red spot level data. 
```{r read-red-spots}
red_spots <- dir_info("data", 
       recurse = T, 
       glob = glob_r) %>%
  filter(size > "1K") %>%
  rowwise() %>%
  summarise(fread(path)) %>%
  select(screen = ScreenName,
         plate = PlateName,
         well = WellName,
         row = Row,
         col = Column,
         field = Field,
         timepoint = Timepoint,
         spot_index = `Object Number`,
         nuclei_index = `Spots 561 - Object No in Nuclei population`,
         x = `Position X [µm]`,
         y = `Position Y [µm]`,
         z = `Spots 561 - Z-plane Mean`) %>%
  ## Assign color label
  mutate(color = "red")

glimpse(red_spots)
```

```{r}
spots <- bind_rows(green_spots, red_spots) %>%
  inner_join(nuclei, by = c("plate", "well", "field",
                          "timepoint", "nuclei_index")) %>%
  select(screen = screen.x,
         plate, well, sgrna,
         row = row.x,
         col = col.x,
         field, timepoint, nuclei_index, nuc_area, 
         n_green_spots, n_red_spots, spot_index, 
         color, x, y, z)
```

Plot the number of green spots per cell, aggregating on sgRNA treatment.
```{r n-green-spots, echo=F}
nuclei %>%
  ggplot(aes(x = n_green_spots,
             y = ..density..)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 0:6) +
  coord_cartesian(xlim = c(-1, 6)) +
  facet_wrap(vars(sgrna)) +
  xlab("Number of Green Spots per Cell") +
  ggtitle("Green Spots") 
```
```{r n-red-spots, echo=F}
nuclei %>%
  ggplot(aes(x = n_red_spots,
             y = ..density..)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = 0:6) +
  coord_cartesian(xlim = c(-1, 6)) +
  facet_wrap(vars(sgrna)) +
  xlab("Number of Red Spots per Cell") +
  ggtitle("Red Spots") 
```
```{r cell-count, echo = FALSE}
nuclei %>%
  group_by(well, sgrna) %>%
  summarise(n_cells = n()) %>%
  ggplot(aes(x = sgrna, y = n_cells)) +
  geom_point() +
  stat_summary(geom = "crossbar",
               fun.data = "mean_sdl",
               fun.args = list(mult = 1))
```



Filter cells that do no haveve 3 spots in both the green and red channel.
```{r}
spots_filtered <- spots %>%
  filter(n_green_spots == 3, n_red_spots == 3) %>%
  data.table()
```

Calculate 2D spot distances.
```{r calc-spot-dist-2D, results="hide"}
#Green-Red
gr_dist_2D <- spots_filtered[, reshape2::melt(dist2(as.matrix(.SD[color == "green",
                                                       list(x,y)]),
                                                    as.matrix(.SD[color == "red",
                                                       list(x,y)])),
                                   value.name = "gr_dist"),
                             by = list(plate,
                                       well,
                                       row, 
                                       col,
                                       sgrna,
                                       field, 
                                       timepoint, 
                                       nuclei_index)]

setnames(gr_dist_2D, c("Var1","Var2"), c("g_index","r_index"))

glimpse(gr_dist_2D)
```

Calculate 3D spot distances.
```{r calc-spot-dist-3D, results="hide"}
#Green-Red
gr_dist_3D <- spots_filtered[, reshape2::melt(dist2(as.matrix(.SD[color == "green",
                                                       list(x,y,z)]),
                                                    as.matrix(.SD[color == "red",
                                                       list(x,y,z)])),
                                   value.name = "gr_dist"),
                             by = list(plate,
                                       well,
                                       row, 
                                       col,
                                       sgrna,
                                       field, 
                                       timepoint, 
                                       nuclei_index)]

setnames(gr_dist_3D, c("Var1","Var2"), c("g_index","r_index"))

glimpse(gr_dist_3D)
```
Green-Red and Red-FarRed proximity calculations. **Important: this calculations find the closest Green/Red  pairs (by calculating the the minimum distance) on a per Red Spot basis**.
```{r proximity-red-green, warning=F}
#Green-Red
setkey(gr_dist_2D, row, col, field, timepoint, nuclei_index, r_index)

gr_dist_min_2D <- gr_dist_2D[, .SD[which.min(gr_dist),], 
                        by = key(gr_dist_2D)]

setkey(gr_dist_3D, row, col, field, timepoint, nuclei_index, r_index)

gr_dist_min_3D <- gr_dist_3D[, .SD[which.min(gr_dist),], 
                        by = key(gr_dist_3D)]
```

```{r sgrna-gr-2D-density, echo=F}
scramble_2D <- gr_dist_min_2D %>%
  filter(sgrna == "SCRAMBLE") %>%
  select(-sgrna)

gr_dist_min_2D %>%
  filter(!sgrna %in% c("PLK1", "SCRAMBLE")) %>%
  ggplot(aes(x = gr_dist)) +
  geom_density(color = "#5DA5DA") +
  geom_density(data = scramble_2D,
               color = "#FAA43A") +
  facet_wrap(vars(sgrna)) +
  coord_cartesian(xlim = c(0, 1.5)) +
  xlab("Green/Red Min. Distance (microns)") +
  ylab("Distance")
```

```{r sgrna-gr-3D-density, echo=F}
scramble_3D <- gr_dist_min_3D %>%
  filter(sgrna == "SCRAMBLE") %>%
  select(-sgrna)

gr_dist_min_3D %>%
  filter(!sgrna %in% c("PLK1", "SCRAMBLE")) %>%
  ggplot(aes(x = gr_dist)) +
  geom_density(color = "#5DA5DA") +
  geom_density(data = scramble_3D,
               color = "#FAA43A") +
  facet_wrap(vars(sgrna)) +
  coord_cartesian(xlim = c(0, 1.5)) +
  xlab("Green/Red Min. Distance (microns)") +
  ylab("Distance")
```

```{r sgrna-gr-3D-histogram, echo = FALSE}
gr_dist_min_3D %>%
  filter(!sgrna %in% c("PLK1", "SCRAMBLE")) %>%
  ggplot(aes(x = gr_dist, y = ..density..)) +
  geom_histogram(color = "#5DA5DA", 
                 binwidth = 0.05,
                 alpha = 0.3) +
  geom_histogram(data = scramble_3D,
                 binwidth = 0.05,
                 color = "#FAA43A",
                 alpha = 0.3) +
  facet_wrap(vars(sgrna)) +
  coord_cartesian(xlim = c(0, 1.5)) +
  xlab("Green/Red Min. Distance (microns)") +
  ylab("Distance")
```


```{r z-histogram, echo = FALSE}
spots_filtered %>%
  ggplot(aes(x = z, y = ..density..)) +
  geom_histogram(binwidth = 0.5,
                 fill = "#5DA5DA",
                 color = "grey30") +
  facet_wrap(vars(color),
             nrow = 2)
```

Document the information about the analysis session
```{r sessionInfo, include=TRUE, echo=TRUE, results="markup"}
sessionInfo()
```